SUPPLEMENT to \\ Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-05-20 doc/mc-paper.tex
Include more experimental results

Computing many useful properties of Boolean formulas, such as their weighted or unweighted model count,
is intractable on general representations. It can become tractable when formulas are expressed in a
special form, such as the deterministic,
decomposable, negation normal form (det-DNNF)\@.
\emph{Knowledge compilation} is the process of converting a formula
into such a form. Unfortunately existing knowledge compilers provide no guarantee that their output correctly
represents the original formula, and therefore they cannot validate a model count, or any other computed value.

@@

We have developed a program that generates POG representations from det-DNNF
graphs 
produced by the state-of-the-art knowledge compiler
\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
are equivalent to the input CNF formulas.  Our toolchain
for generating and verifying POGs scales to all but the largest
graphs produced by \dfour{} for formulas from a recent model counting
competition. Additionally, we have developed a formally verified CPOG
checker and model counter for POGs in the \lean{} proof assistant.
In doing so, we proved the soundness of our proof framework. These programs
comprise the first formally verified toolchain for weighted and unweighted
model counting.
\end{abstract}

@@

One approach to model counting, known as \emph{knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the \emph{deterministic
decomposable, negation normal form}
(det-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004}
represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a det-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
generates a \emph{decision} DNNF (dec-DNNF) representation (a restricted form of det-DNNF),
and the \emph{proof
  generator} uses this to generate a CPOG file.
The \emph{proof checker} verifies the equivalence of the CNF and CPOG representations.
The \emph{ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

@@

In more detail, \cdfour{} emits a sequence of clauses $R$ that
includes the conflict clauses that arose via a top-down processing of
the input clauses.  Given input formula $\phi$, their first task is to
check whether $\phi \imply R$, i.e., that any assignment that satisfies $\phi$ also satisfies each of the clauses in $R$.
They then base other parts of their proof on that
property and use a separate checker to perform a series of additional
checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
satisifies the \emph{reverse asymmetric tautology} (RAT) property with
respect to the preceding
clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
clause $C$ to a set of clauses maintains satisfiability, but it may
not preserve models.  As an example, consider the following formulas:
\begin{center}
  \begin{tabular}{lccc}
    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_1 \lor \obar{x}_2)$\\    
  \end{tabular}
\end{center}
Clearly, these two formulas are not equivalent---$\phi_1$ has six
models, while $\phi_2$ has five.  In particular, $\phi_1$ allows
arbitrary assignments to variable $x_2$.  Critically, however the
second clause of $\phi_2$ is RAT with respect to the first---any
satisfying assignment to $\phi_1$ can be transformed into one that
also satisfies $\phi_2$ by setting $x_2 = 0$.

@@

We summarize the results of our experiments here.  
The supplement~\cite{bryant:sat:2023:supplement}
provides a more complete description. For our
evaluation, we used the public benchmark problems from the 2022
standard and weighted model competitions.  We found that there were
180 unique CNF files among these, ranging in size from 250 to
2,753,207 clauses.
%%We ran our programs on a processor with 64~GB of
%%memory and having an attached high-speed, solid-state
%%disk.
With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We ran our proof generator with a time limit of 10,000 seconds.
The results are shown in Figure~\ref{fig:dual:total}, with the plot of runtimes on the left and the plot of the total number of clauses on the right.
The proof generator was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The prototype checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
Overall the ratio
between the combined generation plus checking time versus the time
for \dfour{} had a harmonic mean of 5.5.
The ratio between the total number of clauses and the number of defining clauses had a harmonic mean of 3.13.
It is worth noting that, to date, we have not found
any errors in the dec-DNNF graphs generated by \dfour{}.

We found that the monolithic approach for generating the forward
implication proof works well for smaller POGs, but it becomes
unreliable and inefficient for larger ones.
%% These experiments suggest
%% a possible hybrid approach, stopping the recursion of the structural
%% approach and shifting to monolithic mode once the subgraph size is below some
%% threshold.
We also found that our two optimizations: literal grouping and
lemmas can provide substantial improvements in proof size and runtime.
Running the verified proof checker in \lean{} showed the same scaling
with respect to the proof size as did the prototype checker, but it
required, on average, around 5.9 times longer.

Finally, the runtimes for the MICE toolchain versus our CPOG toolchain
showed little correlation, reflecting the fact that the two solve
different problems and use different approaches.  In general,
our CPOG toolchain showed better scaling, in part due to its ability
to control the recursion through lemmas.

>>>>

Computing many useful properties of Boolean formulas, such as their weighted or unweighted model count,
is intractable on general representations. It can become tractable when formulas are expressed in a
special form, such as the decision-decomposable, negation normal form (dec-DNNF)\@.
\emph{Knowledge compilation} is the process of converting a formula
into such a form. Unfortunately existing knowledge compilers provide no guarantee that their output correctly
represents the original formula, and therefore they cannot validate a model count, or any other computed value.

@@

We have developed a program that generates POG representations from dec-DNNF
graphs 
produced by the state-of-the-art knowledge compiler
\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
are equivalent to the input CNF formulas.  Our toolchain
for generating and verifying POGs scales to all but the largest
graphs produced by \dfour{} for formulas from a recent model counting
competition. Additionally, we have developed a formally verified CPOG
checker and model counter for POGs in the \lean{} proof assistant.
In doing so, we proved the soundness of our proof framework. These programs
comprise the first formally verified toolchain for weighted and unweighted
model counting.
\end{abstract}

@@

One approach to model counting, known as \emph{knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the \emph{deterministic, decomposable, negation normal form}
(det-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004}, as well as the more restricted
\emph{decision-DNNF} (dec-DNNF)~\cite{huang:jair:2007,oztok:cp:2014},
represent a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a det-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
generates a dec-DNNF representation,
and the \emph{proof
  generator} uses this to generate a CPOG file.
The \emph{proof checker} verifies the equivalence of the CNF and CPOG representations.
The \emph{ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

@@

In more detail, \cdfour{} emits a sequence of clauses $R$ that
includes the conflict clauses that arose via a top-down processing of
the input clauses.  Given input formula $\phi$, their first task is to
check whether $\phi \imply R$, i.e., that any assignment that satisfies $\phi$ also satisfies each of the clauses in $R$.
They then base other parts of their proof on that
property and use a separate checker to perform a series of additional
checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
satisifies the \emph{reverse asymmetric tautology} (RAT) property with
respect to the preceding
clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
clause $C$ to a set of clauses maintains satisfiability, but it may
not preserve models.  As an example, consider the following formulas:
\begin{center}
  \begin{tabular}{lccc}
    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_2 \lor \obar{x}_3)$\\    
  \end{tabular}
\end{center}
Clearly, these two formulas are not equivalent---$\phi_1$ has six
models, while $\phi_2$ has four.  In particular, $\phi_1$ allows
arbitrary assignments to variable $x_2$.  Critically, however the
second clause of $\phi_2$ is RAT with respect to $\phi_1$---any
satisfying assignment to $\phi_1$ can be transformed into one that
also satisfies $\phi_2$ by setting $x_2 = 0$.

@@

We summarize the results of our experiments here.  
The supplement~\cite{bryant:sat:2023:supplement}
provides a more complete description. For our
evaluation, we used the public benchmark problems from the 2022
standard and weighted model competitions\footnote{Downloaded from \url{https://mccompetition.org/2022/mc_description.html}}.  We found that there were
180 unique CNF files among these, ranging in size from 250 to
2,753,207 clauses.
%%We ran our programs on a processor with 64~GB of
%%memory and having an attached high-speed, solid-state
%%disk.
With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We ran our proof generator with a time limit of 10,000 seconds.
The results are shown in Figure~\ref{fig:dual:total}, with the plot of runtimes on the left and the plot of the total number of clauses on the right.
The proof generator was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The prototype checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
Overall the ratio
between the combined generation plus checking time versus the time
for \dfour{} had a harmonic mean of 5.5.
The ratio between the total number of clauses and the number of defining clauses had a harmonic mean of 3.13.
It is worth noting that, to date, we have not found
any errors in the dec-DNNF graphs generated by \dfour{}.

We found that the monolithic approach for generating the forward
implication proof works well for smaller POGs (up to one million
defining clauses), but it becomes inefficient for larger ones.  These
experiments suggest a possible hybrid approach, stopping the recursion
of the structural approach and shifting to monolithic mode once the
subgraph size is below some threshold.  By using monolithic mode, we
were also able to perform end-to-end verification of all but one of the
benchmarks that could be verified without preprocessing, with a total
time limit (including preprocessing) of 1,000 seconds.

We also found that our two optimizations: literal grouping and
lemmas can provide substantial improvements in proof size and runtime.
In the extreme cases, a lack
of lemmas caused one proof to grow by a factor of 52.5, while a lack
of literal grouping caused another proof to grow by a factor of 39.6.
Overall, it is clear that these two optimizations are worthwhile, and
sometimes critical, to success for some benchmarks.

Running the verified proof checker in \lean{} required, on average
(harmonic mean), around 5.9 times longer than the prototype checker.
Encouraging, the scaling trends were identical for the two solvers,
indicating that the two checkers have similar asymptotic performance.
We consider a factor of 5.9 to be acceptable for the assurance
provided by formal verification.

In comparing other proof frameworks, we found that the \cdfour{}
toolchain ran very fast and could handle very large benchmarks.  Even
with a total time limit of 1,000 seconds, including the time for
knowledge compilation, the \cdfour{} toolchain completed for 106
benchmarks, while the CPOG toolchain completed for just 82.  We found
that the runtimes for the MICE toolchain versus our CPOG toolchain
showed little correlation, reflecting the fact that the two solve
different problems and use different approaches.  In general, our CPOG
toolchain showed better scaling, in part due to its ability to control
the recursion through lemmas.  On the other hand, neither of these
toolchains could perform end-to-end verification when the knowledge
compilation was preceded by preprocessing.

### 2023-05-20 doc/mc-supplement.tex
Included plot in main paper
Include more experimental results

As the figures illustrated, each optimization on its own can shorten
the proofs, sometimes substantially, and the two combine to have even
greater effect.  Of the 80 benchmark problems evaluated, and comparing
with both optimizations disabled, 47 had the number of proof clauses
reduced by a factor of 2 by using lemmas, 14 by using literal merging,
and 60 by enabling both optimizations.  In the extreme cases, a lack
of lemmas caused one proof to grow by a factor of 52.5, while a lack
of literal grouping causes another proof to grow by a factor of 39.6.

>>>>

As the figures illustrated, each optimization on its own can shorten
the proofs, sometimes substantially, and the two combine to have even
greater effect.  Of the 80 benchmark problems evaluated, and comparing
with both optimizations disabled, 47 had the number of proof clauses
reduced by a factor of 2 by using lemmas, 14 by using literal merging,
and 60 by enabling both optimizations.  In the extreme cases, a lack
of lemmas caused one proof to grow by a factor of 52.5, while a lack
of literal grouping caused another proof to grow by a factor of 39.6.
